\documentclass{lab}

\usepackage{graphicx}
\usepackage{float}
\usepackage{soul}
\usepackage{multicol}

\addtolength{\oddsidemargin}{-.4in}
\addtolength{\evensidemargin}{-.4in}

\title{ENGG1003 - Lab Week 5}
\author{Brenton Schulz}
\date{\today}

\begin{document}
\maketitle

\section{Introduction}

This lab covers the three topics which will be assessed during the Week 9 assessed lab. It includes:

\begin{itemize}
	\item Functions which accept pointer arguments
	\item Multi-Dimensional arrays
	\item File I/O
\end{itemize}

\section{Pointers}

\begin{task}{}{}
(Basic ``toy'' example)
\\~\\
Write a C function which takes a single \texttt{int *} argument and zeros it. The function prototype is:
\begin{lstlisting}[style=Ctable]
void zeroInt(int *x);
\end{lstlisting}

Write your own ``test code'' to verify the function's operation.
\end{task}

\begin{task}{}{}
Write a C function which accepts two arguments of type \texttt{int *} and swaps them. The function prototype is:

\begin{lstlisting}[style=Ctable]
void swap(int *a, int *b);
\end{lstlisting}

Write your own ``test code'' to verify the function's operation.
\end{task}

\begin{task}{}{}
Write a C function which accepts three \texttt{int *} arguments, allocates the mean of the three numbers to the first argument and zeros the others. The function prototype is:

\begin{lstlisting}[style=Ctable]
void mean(int *a, int *b, int *c);
\end{lstlisting}

Write your own ``test code'' to verify the function's operation.
\end{task}

\pagebreak

\section{Multi-Dimensional Arrays}

\begin{task}{}{}
The template below, write a C program which calculates the mean of the initialised 2D array and prints the result to the console.

\begin{lstlisting}[style=Ctable]
#include <stdio.h>

main() {
	float myArray[3][3] = { { 0.1, 0.2, 0.3 }, { 1.1, 1.2, 1.3 }, { 2.1, 2.2, 2.3 } };
}
\end{lstlisting}
\end{task}


\section{File I/O}

\begin{task}{}{}
Write a C program which reads initialises a 2D array from a file. The file's contents is:

\begin{lstlisting}[style=Ctable]
12 31 35
23 5 43
434 63 64
\end{lstlisting}

Create a file and copy the contents above into it.\\~\\

After reading the data, find the largest number in the 2D array and print its value and indices (ie: location or address within the array).

You may use the following template:

\begin{lstlisting}[style=Ctable]
#include <stdio.h>

main() {
	int arrayData[3][3];
	FILE *input;
	
	// Open the file
	
	// Read the data from the file into arrayData
	
	// Find the largest value, print it, and it's indices
}
\end{lstlisting}
\end{task}

\end{document}
